\chapter{Conclusion}
\label{concl}

\section{Summary}
\label{ssummary}

Network zoning has long been recognized as the cornerstone of secure network operation and management.
In the current practice, operators realize network zones with network segmentation technologies and security middleboxes.
As information systems become more dynamic from a topological, operational, and functional perspective, however, the conventional network-zoning architectures face new challenges in terms of scalability and flexibility.
In this paper, we have shown that lightweight policy enforcement for inter-zone communication
is achievable.
Following a constructive approach with a cryptographic foundation, it is possible to create a proactive alternative to the mostly reactive systems presently used in network zoning.
In conjunction with \name, verification based on firewalls becomes simpler because firewalls would only process a limited amount of (filtered) traffic.
\name consequently reduces the number of management points of distributed networks while retaining a high degree of security.

\section{Future Work}
\label{sfuture}

The work presented in this thesis can be extended along different axes. For
one, it would be interesting to explore how an architecture like \name can be
used to provide privacy for inter-domain data transmissions. \name provides
confidentiality and integrity on the data in transit. However, certain traffic
patterns and packet sizes could still allow an attacker to draw reliable conclusions about
what sort of traffic is being sent (\eg sensor data, control messages, file
transmissions, ...) and which security zones are located
at a given site. If privacy is a concern, a first step could be to extend \tps
with new modules that pad all packets to the full MTU length and insert mock
traffic, such that traffic patterns are hidden.

Another direction in which this work could be extended is the implementation
of the \tp gateway in a high-performance language. Various operations---in
particular cryptographic---would benefit greatly from a state-of-the-art
implementation on dedicated network hardware.

Lastly, it could be interesting to explore how certain properties of local
networks can be conveyed across the WAN to a remote network. For example, technologies such as
SPB and TRILL have support for layer 2 multipathing, the corresponding control
messages are, however, not transmitted by \name. A deeper interoperability of
different protocols across network boundaries would certainly be desirable property.
