\chapter{Introduction}
\label{intro}

% about network security zoning
Network zoning has long been an essential part of the Internet security infrastructure, 
which logically partitions network and information assets into disjoint segments that share the 
same security requirements and policies, and functional similarities. Zones define the network boundaries and their defense 
requirements by stating the entities populating the zones, the entry points into the zone, and
how traffic is monitored and filtered at these entry points. Informally, these zones 
are realized by a virtualized separation at layer 2 (e.g., IEEE 802.1q~\cite{ieee2018vlan}) 
with firewalls at higher levels governing data transfers between zones~\cite{mayer2000fang}. 
% In general, security zones are 

% new requirements for the modern network zoning (motivation)
% > Datacenters zoning (for tenant)
% > Enterprise leveraging LaaS

% Most enterprise networks have embraced the notion of security layered network classification,
% that can be broadly classified as; a restricted network which houses business-critical IT 
% assets, and a public access network where customer services are implemented, mediating access
% from the public Internet. 
% Most enterprise networks comprise with a large number of zones that define operational, 
% organizational, and most importantly security factors.
Each zone is identified by a distinct level of trust, and 
% each pair of zones forms a trusted-untrusted relationship~\cite{obregon2015infrastructure}.
forms a trusted/untrusted relationship with other zones~\cite{obregon2015infrastructure}.
% \claude{there can also be trust-trust and untrust-untrust?}
To realize the unidirectional trust model, firewalls are considered to be the most viable
technology and are widely used in the current practice. However, operating firewalls in 
large enterprises is often challenging for network operators and security architects. The 
access control for network zones might be dynamic, and thus it requires complex 
management schemes to accommodate a myriad of policies. While there are advanced 
technologies such as virtual firewalls~\cite{deng2015vnguard,bakker2016network}, distributed
security enforcement~\cite{markham2001security,yu2017psi}, and Unified
Threat Management (UTM)~\cite{qi2007towards}, newly designed to enforce access control polices in extremely dynamic networks, network zone management and modeling 
still remains cumbersome~\cite{ramasamy2011towards,gontarczyk2015blueprint}.

% limitations on the current zoning techonologies
% > secure communication for inter-domain zones (vxlan, ipsec)
% > cost inefficiency (expensive lease line and firewalls)
% > management scalability (key management)
Bridging geographically distant network zones is very challenging today. In general,
network zones are created not only for security purposes but also because of geographical,
operational, or organizational factors. Large enterprises with geographically distributed 
branch networks, and possibly collaborative partners' networks need to be interconnected. 
Given that distant network zones exchange information over an untrusted 
network (e.g., the Internet), there is a risk that the communication exposes security-sensitive 
information during transit. To mitigate
such threats, administrators leverage additional security mechanisms (e.g.,
IPsec~\cite{rfc4301} and SSL-VPN~\cite{sun2011advantages}) 
% \ml{`VPN' hasn't been introduced yet.} \ml{``IPsec'' is normally written like this (with lower-case `s')} 
which ensure confidentiality
and integrity of the transmission over the untrusted network by encrypting and authenticating the data with
securely shared cryptographic keys. Nonetheless, these technologies bring forth new challenges
such as management scalability~\cite{felsch2018dangers} and compatibility issues with other   
security solutions~\cite{liu2008collaborative}---universal agreement with business partners on building collaborative security infrastructure is often problematic.
% \claude{This is espcially problematic when collaborating with business partners as there is no universal agreement on what solution to use.}

% the notion of distributed data centers

% Passport summary
% > secure zone transfer over wan
% > simplified zoning architecture (logical transit zone)
%   > cost efficient: firewall x, lease line x
% > easy key management (management scalability)
\name is a new network zoning architecture that secures inter-zone communication---which 
operates on layer 3, supporting heterogeneous layer 2 architectures---while ensuring 
scalable cryptographic-key management and flexible security policy enforcement. 
% \ml{TODO: use consistent capitalization of ``layer'' (currently mix of lower and upper case).} 
\name flattens the current hierarchically-complex network zone topology into a collection of 
horizontal zones connected to a unified security gateway, called Zone Translation Point 
(\tp), thus simplifying large enterprise networks.
By interconnecting zones through \tps, complex zone restructuring operations become
easier with respect to new zone initializations or zone migrations.
The \tp ensures source authentication, zone transfer authorization, and illegitimate access filtering by acting as a secure ingress/egress point for network zones. A logically centralized control unit provides management scalability on zone classification and policy enforcement, and mediates cryptographic key establishment. 
% \claude{controller is actually not involved in key managment}
% \jk{Theoretically, it is in a way that the controller issues certificates for \tps and resets the 
% local secret values by sending a key refresh command (although the parts are not implemented).}

% \ml{If the term/abbreviation `\tp' is introduced, it should be used consistently. It is not quite clear if `\tp' and `gateway' refer to the same thing.}
% \ml{In general, all abbreviations should be introduced once at first use and afterwards used consistently (\eg, VLAN)}

A secure zone transfer is performed in three steps: i) the security gateway acquires
access policies for each network zone from its controller, ii) the gateway issues a
cryptographically protected authorization token if a given zone transfer request is 
permitted, and iii) the network forwards only packets with a valid token. By leveraging 
the notion of secure tunneling between two endpoints (i.e., a pair of local and remote \tps), 
confidentiality and integrity of the zone transfer packets are ensured, while keeping the overhead of the authentication process small. For scalable key management, 
we employ a key establishment system that enables dynamic key derivation 
and ensures perfect forward secrecy.

% Evaluation
We provide an implementation of \name that ensures secure zone transfer for
both intra and inter-domain communication at line rate, while requiring no network-stack 
changes from end hosts. We extensively evaluate this implementation to demonstrate
the practical viability of \name. The results show that the \tp introduces negligible 
processing delay; less than \SI{500}{ns} of additional delay for intra-domain zone transfer
and approximately $2.5 \sim 3 \mu$s for inter-domain zone transfer traffic. We further 
provide in-depth analyses for security and practical considerations.

% contributions
% > introduce a new security architecture that enables secure, flexible, practical zoning
% > network filtering at the edge of network
% > provide poc implementation and evaluation
% > provide practical considerations
The main contributions of this paper are the following:

\begin{itemize}
	\item We introduce \name, a new security architecture that enables secure, flexible 
	and viable network zoning and inter-zone communication for large enterprise 
	networks.
	% \claude{while requiring little operational resources}.
	\item We introduce the new notion of an inter-domain transit zone that dramatically 
	simplifies the current hierarchical zone structure, enabling flexible and cost-efficient 
	network zone management. 
	\item \name enables network filtering at the edge of the network, such that it suppresses
	network overhead in terms of wasted network bandwidth and packet processing time.
	\item We implement \name as an opensource project. %that allows anyone to try it out.
	% \item We provide thorough evaluation and analysis including practical considerations.
\end{itemize}

