\chapter{Practical Considerations}
\label{practical}

In this section, we discuss some practical considerations including functional and
management aspects, along with various deployment scenarios describing how \name can
be realized on today's enterprise infrastructure.
% We further discuss about the distributed controller in Appendix~\ref{ssec:distributedcontroller}.


\section{NAT Devices}
\label{sec:nat}
Multiple hosts connected through a NAT device appear as a single host to the
\tp. \textit{Internal} NAT devices hardly pose a problem since the hosts under that NAT
device are all subject to the same subnet and therefore belong to a single network zone.
Network operators establish a security policy on the translated IP address, such
that \tps are able to authenticate the zone transit requests from/to a host under the
internal NAT device.

However, an \textit{external} NAT device located in an external network, e.g., carrier grade NAT,
could affect the \tp's secure tunneling ability. The translated \tp's IP address
would cause a MAC verification failure---recall that each symmetric key binds to the
triplet including \tps' IP addresses as described in \S\ref{sec:keymanagement}.
This, however, can be addressed by enforcing \tps to use their public IP addresses
to derive the symmetric keys. The keys are still secure since the first-level key
from which the pairwise keys are derived is exchanged with the CA-certified public
keys. Discovering the translated \tp address is also not a problem thanks to
the controller informing senders about the recipient \tp's address (see, Protocol 3 in
Figure~\ref{fig:protocol}).

A potential operational failure is where multiple \tps are behind the same NAT device.
This would lead to \tps having identical keys from the view of remote \tps. One possible solution
would be to use a unique \tp identifier instead. Since all such \tps would be under one administration domain, assigning unique \tp identifiers upon bootstrapping is feasible. Then, the \tps convey their identifiers in the AT header field alongside the destination zone ID. This
might slightly increase the size of the header, but does not degrade the security of
the underlying authentication.


\section{Tunneling Granularity} %\paragraph{Zone-to-Zone VPN}
\label{sec:granularity}
% Thanks to the flexible and scalable key derivation scheme introduced in~\cite{rot2020piskes},
% the key establishment supports the secure tunneling in different granularities: 
% i) site-to-site tunneling, ii) zone-to-zone tunneling, and iii) site-to-zone tunneling 
% (the one we take). 
Secure tunneling can be realized in different granularities: i) site-to-site tunneling,
ii) zone-to-zone tunneling, and iii) site-to-zone tunneling (the one used by \name).

% Similar to IPSec VPN, the site-to-site tunneling requires a single symmetric key per
% a \tp pair, e.g., the first-level key. Indeed, the site-to-site key establishment is 
% already enough in terms of communication security and privacy for two tunnel endpoints. 
% However, this approach requires another step of zone transfer verification.
Similar to IPSec VPN, site-to-site tunneling provides strong guarantees about communication
security and privacy for two tunnel endpoints.
However, from a flexibility and manageability standpoint, having a site-to-site tunneling
architecture is not ideal. Every tunnel endpoint needs to share a key with every other
endpoint with which it wishes to exchange data. This adds state to the endpoints that
needs to be kept in sync. Adding a new site requires an update on all the other sites that
wish to communicate with the new site. Then, yet another layer of security middleboxes (e.g., firewalls) are required to perform zone transfer authentication since keys are do not designate a specific zone.

An alternative way of providing authentication is to use one key per zone. In this model,
the \tp would sign the data on behalf of the zones. In case of a zone transfer, the \tp
would do the transfer and then use the key of the source/receiver zone pair. This approach
has the benefit that sender and receiver get decoupled as in principle any site that
contains a given zone is able to decrypt data destined for that zone. Adding a new site
would be as easy as fetching the right keys for the zones used in this site. This process
is independent of all the other sites. However, a receiver \tp needs to be able to fetch
the right keys for the zones, which means the zone transfer information must be visible,
and thus attacker could potentially learn the zone structure of the observed network.
% \claude{only the destination zone id would be leaked, same as in our model} 
Also, having a separate key per zone pair does not scale since the number of zones get big
very quickly for large networks.

Driven by these considerations, we designed the new concept of site-to-zone tunneling,
which represents a middle ground combining the advantages of the two approaches, the
notion of secure tunneling and zone transfer authentication. The symmetric keys
are distinguishable depending on the destination zone, while at the same time the
zone-to-zone security policies are not being exposed. Thanks to the flexible and scalable key derivation scheme introduced in~\cite{rot2020piskes}, the key establishment does not expand
state, while still providing unique symmetric keys per zone.


% \begin{table}
%   \caption{Comparison of nonce reset strategies.}
%   \label{tab:comparison}
%   \setlength{\tabcolsep}{4pt}
%   \newcommand{\rot}[1]{\multicolumn{1}{c}{\adjustbox{angle=60,lap=\width-1em}{#1}}}
%   \newcommand{\rotLast}[1]{\multicolumn{1}{c}{\adjustbox{angle=60}{#1}}}
%   \centering
%   \resizebox{0.485\textwidth}{!}{
%     \begin{tabular}{*6{r}}
%       \toprule
%       % & \rot{Requires randomness} & \rot{Requires non-volatile memory} & \rot{Overlap probability (no reset)} & \rot{Overlap probability (R resets)} & \rot{Overlap} \\ 
%       \midrule
%       Counter only   & \xmark          & \xmark        & 0       & 1     & full \\
%       Randomness only   & \cmark          & \xmark        & $1-(N!/(N^R-(N-R)!))$       & $1-(N!/(N^R-(N-R)!))$    & partial \\
%       Counter paired with randomness   & \cmark          & \xmark        & 0       & $1-(S!/(S^R-(S-R)!))$     & full \\
%       Reset points   & \xmark          & \cmark        & 0       & 0     & - \\ 
%       \bottomrule
%     \end{tabular}
%   }
% \end{table}


\section{Distributed Controllers}
\label{sec:distributedcontroller}
Driven by the scalability and reliability issues inherent to a single physical controller,
namely \textit{single-point-of-failure}, logically centralized control-planes built on
physically distributed instances find wide acceptance in the current practice. The most
common approaches realizing distributed controllers can be broadly categorized into
horizontal distribution~\cite{berde2014onos,medved2014opendaylight} and hierarchical
distribution~\cite{hassas2012kandoo,yap2017taking}. Independent of which distribution
architecture is used, we discuss location, coordination, and migration aspects of distributed
controllers.

\paragraph{Location}
The notion of a logically centralized control-plane offers flexibility in network design
and management. A key design choice is placement of the (distributed) controllers, which
could impact to performance, reliability, and management scalability of a given network.
There is comprehensive research on the controller placement problem considering practical
issues from control latency to reliability, from cost-optimization to load balancing, and so
on~\cite{das2019survey,zhang2017role,he2019toward}. Among those, we are mainly interested in the
latency performance indicator; that is the latency between a controller and regional forwarding
devices.

The best latency is achieved when each branch site has its own controller. By a placement
near local \tps, the controller minimizes the \tp-controller latency for the zone
transfer authorization protocol, allowing instant feedbacks for packet forwarding---we note
that inter-controller communication for global coordination is commonly not latency sensitive.
For the sake of control-plane security, the controller resides in a highly restricted zone to
which only the local \tps and remote controllers have access. Although the per-site controller
offers the best performance regarding policy enforcement for the data-plane, there might be a
cost-efficiency problem for a large-scale network with thousands of branches.

Alternatively, we consider a sparse distribution model, e.g., on edge-cloud systems.
Similar to today's cloud services, network operators running geographically distributed
data centers can instantiate multiple controllers at the central point of regional branches.
% \claude{branch centric?} 
The control-plane latency overhead would be relatively higher compared to dense deployment
model---if the data center edges are geographically diverse, the overhead could be
minimized---but, in terms of cost-optimization and management scalability, it could
be a more viable approach.

\paragraph{Coordination}
It is important to keep consistency in global coordination across the distributed controllers.
Indeed, inconsistency in security policy might grant hosts with a low security
clearance unauthorized access to highly restricted zones, resulting in unexpected information
leakage and eventually administration failure. With this in mind, we consider a consensus
algorithm with strong consistency guarantees~\cite{panda2013cap,phemius2014disco,shi2014giraffe},
where the security policy is dynamically shared/replicated across the distributed controller
instances, ensuring concurrent policy enforcements toward the data-plane devices. There
are numerous open-source projects, such as \fnurl{Consul}{https://github.com/hashicorp/consul},
\fnurl{Apache ZooKeeper}{https://zookeeper.apache.org/}, and \fnurl{ETCD}{https://github.com/etcd-io/etcd} available.

\paragraph{\tp Migration}
To benefit from the distributed controller environment, a dynamic controller discovery
process also becomes important. That is, \tps should be able to search a cluster of best
candidates, diagnose the performances of control latency, and seamlessly migrate to the
best controller. To this end, we consider a two-step migration process: i) \tp-driven
control channel initialization and ii) controller-driven \tp migration.

\tps are responsible for establishing the first control plane channel with a controller.
For example, a new \tp ($\tp_{new}$) has been configured to contact an initial controller
acting as a first rendezvous point.
% \claude{a tp has been configured with an initial controller address acting as rendez-vous point?}
The initial information contains the controller's IP address
($C$), the corresponding zone ID ($Z_{C}$), and the \tp's IP address behind which the controller
resides ($\tp_{C}$). If the controller locates in the remote site (i.e., $\tp_{new}
\neq \tp_{C}$), $\tp_{new}$ should connect with $C$ through $\tp_{C}$. Otherwise, e.g.,
$C$ is within the same LAN or in public network, $\tp_{new}$ can directly send $C$ a request
for control-plane channel establishment.

Once the \tp joined the network, the controller then initiates a migration process to
find the best controller ($C_{best}$) for $\tp_{new}$. Upon a migration request broadcasted
by $C$, other controllers measure the possible latency to $\tp_{new}$ and reply back
the results. Then, $C$ elects $C_{best}$ considering the latency measurements and the
current load balance, and sends $\tp_{new}$ a \texttt{RoleChange()} request containing
$C_{best}$, $Z_{C_{best}}$, and $\tp_{C_{best}}$. Finally, $\tp_{new}$ swaps the best
controller by establishing a new channel with $C_{best}$. The migration process
is also applied when changes in the network are detected.


\section{Nonce Reset}
\label{sec:nonce}
% Authenticators are created using an AEAD algorithm with an underlying AES-128 cipher in GCM mode. GCM requires a nonce as input argument to build counters which are then used to create the keystream. 
The same nonce must never be used twice with the same key, otherwise the security of the
cipher significantly decreases. In theory it is easy to create nonces that fulfill this
requirement. One can simply use a counter which is increased for every invocation of the
AEAD algorithm. In real systems this is not so easy to achieve since machines can crash
and lose their state, specifically their nonce counter. Outlined below are some techniques
to approach this problem.

\begin{itemize}
	\item \textit{Purely random nonce}: uses all bits of the nonce for randomness. Low
	      probability for overlap but no guarantee even when no resets.
	\item \textit{Counter paired with random sequence}: divides the nonce into a counter
	      and a randomized part. Initialize the random part after every restart and increase
	      the counter part for every packet. On reset start counter from zero with a fresh
	      randomized part. In case of overlap all packets do overlap.
	\item \textit{Reset points}: defines specific resets points for the counter which are
	      stored on non-volatile memory (NV-memory). Increment counter in memory and write the
	      next reset point to NV-memory when threshold is crossed. On crash restart counter from
	      reset point on NV-memory.
\end{itemize}


% \subsection{DMZ Traffic}
% So far, we mainly described on how two \tps can securely communicate each other. However,
% since most of cooperative information systems have public services facing the public Internet,
% we also need to consider 

% DMZ is a zone hosting public services. Since access traffic to the public services is 
% mostly originated from public Internet, 

% \subsection{Fast Failover}
% \label{ssec:ecmp}
% Operating multiple \tps with advanced ECMP-enabled layer 2 protocols (e.g., SPB and 
% TRILL) is a viable network design that provides load balancing and enhanced resiliency 
% against a \tp or link failure, as we discussed in \S\ref{ssec:disrupting}. 
% If a \tp is unable to continue data transmission, ECMP engages. It searches an alternative
% forwarding path considering cost equality and redirects flows through the new path, assuring
% continuous communication for end hosts~\cite{rfc6754}. 
% This elastic migration during a transient network outage, however, does not replicate the
% forwarding state of each flow, and thus might require another step of repopulating  
% forwarding state. Upon packet arrival, the new \tp fetches the zone transfer policy
% and caches in the forwarding table. Note that, for \texttt{established}, it might
% cause unexpected packet drops if the first packet is a response. Nevertheless, the 
% communication would continue as soon as the original sender noticed the packet drop 
% and requests retransmission. 


% \subsection{Cross-enterprise Setup}
% \label{ssec:crossconnection}
% Collaborators may join to an enterprise network. To be able to access highly secure zones
% in the enterprise network, the collaborators need to satisfy the following administrational
% and technical constraints: i) the collaborators must acquire a proper access grant from
% the enterprise's administration, and ii) install a \tp at their network.

% The enterprise network would have a single administration domain as we assumed in~\S\ref{ssec:assumptions}. To allow the collaborators interconnect their network system

% \subsection{\tp Deployment}
% \label{ssec:deployment}

% From a technical perspective, a \tp has obligations of: i) interconnecting VLANs , ii) 
% fetching security policies from the controller, and iii) enforcing the policies. 


% \paragraph{Gateway}
% A gateway 


% \paragraph{Middlebox}



\section{Incremental Deployability}
\label{sec:deployability}

To be incrementally deployable, \name does not require changes from end hosts nor the local network infrastructure. \name can take a supportive role by complementing already installed lines of defense such as firewalls, IPS, and IDS. For instance, traffic can be pre-filtered by \tps before it reaches firewalls located deeper inside the network. This approach favors an incremental deployment strategy. On the other hand, \name can also be used as a single, all-in-one solution providing packet filtering, tunneling, and routing within one device. Such a deployment is especially interesting for small branch sites with much simpler network layouts. Here, \name can drastically reduce the number of devices that need to be maintained.



