\chapter{Packet Metadata}
\label{apdx:meta}

%\section{Packet Metadata}
%\label{apdx:metadata}

\begin{lstlisting}[
	basicstyle=\footnotesize, language=Go,
]
type Packet struct {
	Ingress		bool
	SrcHost 	net.IP
	DstHost 	net.IP
	RemoteTP	string
	DstZone		uint32
	RawPacket 	common.RawBytes
}
\end{lstlisting}

The packet metadata is the abstract object passed between modules, describing an IP packet. 
It accumulates information about the raw IP packet (\texttt{RawPacket}). The \texttt{Ingress} 
field identifies a packet as either an ingress packet, coming from the WAN, or an egress packet 
that originated in the local network. \texttt{SrcHost} and \texttt{DstHost} reflect the source 
and destination IP addresses of the packet. \texttt{RemoteTP} designates the remote \tp. For an 
ingress packet that is the source \tp from which the packet was received, for an egress packet 
it is the \tp to which the packet needs to be forwarded to. \texttt{DstZone} is the Zone ID of 
the zone to which \texttt{DstHost} belongs.

\chapter{Controller Database}
\label{apdx:controllerdb}


\begin{lstlisting}[language=sql, basicstyle=\tiny, %or \small or \footnotesize etc.
]
CREATE TABLE Zones(
id INTEGER NOT NULL,
name TEXT,
PRIMARY KEY(id)
);

CREATE TABLE Sites(
tp_address TEXT NOT NULL,
name TEXT,
PRIMARY KEY(tp_address)
);
	  
CREATE TABLE Subnets(
net_ip BLOB NOT NULL,
net_mask BLOB NOT NULL,
zone INTEGER NOT NULL,
tp_address TEXT NOT NULL,
PRIMARY KEY (net_ip, net_mask),
FOREIGN KEY (zone) REFERENCES Zones(id) ON DELETE CASCADE,
FOREIGN KEY (tp_address) REFERENCES Sites(tp_address) ON DELETE CASCADE
);
	  
CREATE TABLE Transfers(
src INTEGER NOT NULL,
dest INTEGER NOT NULL,
PRIMARY KEY (src, dest) ON CONFLICT REPLACE,
FOREIGN KEY (src) REFERENCES Zones(id) ON DELETE CASCADE,
FOREIGN KEY (dest) REFERENCES Zones(id) ON DELETE CASCADE	
)
\end{lstlisting}

The controller database consists of 4 tables: \texttt{Zones}, \texttt{Sites}, \texttt{Subnets}, 
and \texttt{Transfers}. The \texttt{Zones} table contains all network zones known to the 
controller, identified by zone IDs. Additionally, a human readable description is attached. The 
\texttt{Sites} table holds all known branch sites with the addresses of the corresponding \tps 
and a textual description. The \texttt{Subnets} table describes the configured IP subnets 
together with their zone membership and the \tp behind which they are located. Finally, the 
\texttt{Transfers} table reflects the zone transfer matrix of allowed zone transfers.